---
title: "Gov 50 Final Project"
author: "Kareem Fahs"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

Do American citizens vote less in presidential elections during times of economic success and more in times of economic decline? In this study, I plan to examine the correlation between voter turnout and GDP in presidential election years. I hypothesize that in times of economic decline, citizens will be more likely to vote in an effort to drive change in economic policy. Concurrently, I believe that in times of relative economic prosperity voters will experience a sense of security and maybe not be as motivated to go to the polls. I would combine two separate data sets for this study: one on the turnout results of every state in every presidential election since 1976, and the other on the GDP of the U.S. in every year from 1960-2023. The explanatory variables would be the yearly GDP and GDP per capita of the U.S, with the outcome variable being the turnout in presidential elections since 1976. If we see a correlation in which turnout (X) is typically higher when GDP (Y) is low, that would support my hypothesis that low economic strength leads to the electorate opting for change. If instead, the data reflects relatively steady turnout rates that are not affected by GDP that may contribute to disproving my hypothesis. Another thing that could disprove my hypothesis is if we actually find that turnout is higher during times of economic strength; this could be explained by the electorate being motivated to protect the strength of the economy leading to higher turnout. 




```{r}
#Loading Packages
library(tidyverse)
library(infer)
library(scales)
library(readr)
```

```{r}
# loading data 
turnout <- read_csv("~/Downloads/VEP_Turnout.csv", show_col_types = FALSE)

gdp <- read_csv("~/Downloads/us_gdp.csv", show_col_types = FALSE)

turnout
gdp
```



```{r}
# Pivoting GDP Data to make Years rows

gdp_long <- gdp |>
  pivot_longer(
    cols = `1960`:`2020`,
    names_to = "year",
    values_to = "GDP"
  ) |>
  mutate(year = as.integer(year))

gdp_long


```

```{r}
#Joining the Data

gdp_turnout <-gdp_long |>
  cross_join(turnout)

gdp_turnout_presidential <- gdp_turnout |>
  subset(year == year_presidential)

gdp_turnout

gdp_turnout_presidential
```

```{r}
gdp_turnout |>
  ggplot(aes(x = year, y = GDP)) + 
  geom_point() +
  labs(x = "Year", y = "GDP", title = "U.S. GDP by Year") +
  scale_y_continuous(
    labels  = label_number(scale = 1e-12, 
                           prefix = "$", 
                           suffix = "t", 
                           accuracy = 1)
    ) 
```



```{r}
# Plotting GDP against Turnout
library(broom)



gdp_turnout_presidential |>
  group_by(year_presidential) |>
  ggplot(aes(x = GDP, y = vep_presidential_turnout)) + 
  geom_point() +
  geom_smooth() +
  labs(title = "Presidential Turnout Based on GDP") +
  scale_x_continuous(
    labels  = label_number(scale = 1e-12, 
                           prefix = "$", 
                           suffix = "t", 
                           accuracy = 1))

fit_1 <- lm(vep_presidential_turnout ~ GDP, data = gdp_turnout_presidential)

fit_1


```


  Based on the results of this second visualization, we can see that any particular GDP can have any outcome of turnout. This trend holds true for both times of economic decline and times of economic success. With that being said, based on the plot there seems to be a slight relation between GDP and turnout in presidential elections and this is backed up by the linear regression. The linear model suggests that for a one unit increase in GDP there is a near zero increase in Presidential voter turnout.  When analyzing the graph, it is important to note that there are more elections that occurred during times of relative economic decline but this may be simply due to the fact that they represent multiple decades prior to today. These low turnouts for low GDPs may be counteracting high turnouts for high GDPs, creating a near zero relation. This does not completely rule out my hypothesis because within a shortened time frame, it is possible that we will see more of a relation between GDP and voter turnout.










